Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-04-20 doc/invt.tex
invt.tex: edits

Another special case is the
one-dimensional one-step process\cite{vankampen},
which models a local sampling process.
In each step, the system hops to
one of the two neighboring bins
with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf Q
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      \frac {1 - \sigma} 2 & \dots &  & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For many updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf Q$ and $\mathbf w$
share the eigenvectors, but not the eigenvalues.
%
The eigenvalues of $\mathbf Q$ are
$$
\Lambda_k
=
1 -
2 \sin^2 \frac{ (k - 1) \, g \, \pi }
              {       2 \, n        }
,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with $\kappa_k(0) = 1$ being the same as
the value in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
  \Gamma_k
  =
  1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
  =
  \frac{ 1 + \Lambda_k }
       { 1 - \Lambda_k }
  =
  \cot^2 \left[
    \frac{ (k - 1) \, g \, \pi }
         { 2 \, n }
  \right]
  .
\label{eq:Gamma_onestep}
\end{equation}
%

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
%\label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  ,
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically
through the normalized mass distribution
along $q$ as
\begin{equation}
  m(q)
  =
  \frac{
    \sqrt{
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2(q)
    }
  }
  {
    \int_0^{ q(T) }
      \sqrt{
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2(q')
      }
      \, d q'
  }
  ,
\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(q) = 1$.
%
As shown in Fig. ,
the distribution peaks at $q = q(T)$,
and decreases with the difference $q(T) - q$.
%
According to Eq. \eqref{eq:q_opt},
the ordinate of the distribution curve at $q$
gives $1/(T\alpha) = 1/(T \dot q)$,
and the area under the curve in the region $q' < q$
gives $t/T$.
%
From this picture, we can draw several conclusions.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  }
  \caption{
    \label{fig:massq}
    Geometric characterization of the optimal schedule,
    $\alpha(t)$,
    via the normalized mass distribution, $m(q)$.
    %
    \note{The figure was produced by \texttt{doc/fig/massq.gp}.
    }%
  }
\end{center}
\end{figure}

@@

  \node[draw, text width={width("System and")}]
    (sys) at (0, 1) {System and\\sampling \\process};

  \node[draw, text width={width("Eigenvalues, ")}]
    (lambda) at (2.5, 3) {Eigenvalues, \\$\lambda_2, \dots, \lambda_n$};

  \node[draw, text width={width("Integrals, ")}]
    (gamma) at (2.5, 1) {Integrals, \\$\Gamma_2, \dots, \Gamma_n$};

  \node[draw, text width={width("Simulation ")}]
    (T) at (2.5, -0.5) {Simulation \\time, $T$};

  \node[draw, text width={width("Schedule ")}]
    (alpha) at (5.2, 1) {Optimal schedule, \\ $\alpha(t)$};

  \node[draw, text width={width("Error ")}]
    (err) at (7, 1) {Error, \\ $\Err$};

  \node[inner sep=0, minimum size=0] (M1) at (3.7, 1) {};
  \node[inner sep=0, minimum size=0] (M2) at (4.0, 1) {};
  \node[inner sep=0, minimum size=0] (N2) at (4.0, -0.2) {};
  \node[inner sep=0, minimum size=0] (N3) at (6.2, -0.2) {};
  \node[inner sep=0, minimum size=0] (M3) at (6.2, 1) {};

  \draw[->, >=latex, thick]
    (w) edge (lambda)
    (w) edge (gamma)
    (sys) edge (gamma)
    (alpha) edge (err);

  \draw[->, >=latex, thick]
    (lambda) -| (M1)
    (gamma) -- (M1)
    (T) -| (M1)
    (M1) -- (alpha);

  \draw[->, >=latex, thick]
    (M2) -- (N2) -- (N3) -- (M3);

\end{tikzpicture}
\caption{
  \label{fig:vardep}
  Dependence of variables in computing
  the optimal schedule and the error.
}
\end{figure}

@@

\item
Select an initial value of $q(T)$,
e.g., $\ln(1 + \alpha_0 \, T)$.

@@

\item
Compute the error $E$ from
Eqs. \eqref{eq:error_split},
  \eqref{eq:error_asym1} and
  \eqref{eq:error_res1}.

\item
Modify the value of $q(T)$ and repeat the previous two steps,
until the minimal error $E$ is found.

>>>>

Another special case is the
one-dimensional one-step process\cite{vankampen},
which models a local sampling process.
In each step, the system hops to
one of the two neighboring bins
with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf A
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      \frac {1 - \sigma} 2 & \dots &  & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For many updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf A$ and $\mathbf w$
share the eigenvectors, but not the eigenvalues.
%
The eigenvalues of $\mathbf A$ are
$$
\Lambda_k
=
1 -
2 \sin^2 \frac{ (k - 1) \, g \, \pi }
              {       2 \, n        }
,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with $\kappa_k(0) = 1$ being the same as
the value in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
  \Gamma_k
  =
  1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
  =
  \frac{ 1 + \Lambda_k }
       { 1 - \Lambda_k }
  =
  \cot^2 \left[
    \frac{ (k - 1) \, g \, \pi }
         { 2 \, n }
  \right]
  .
\label{eq:Gamma_onestep}
\end{equation}
%

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  ,
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define a mass function
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and a normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  ,
\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
As shown in Fig. \ref{fig:massq},
the distribution peaks at $Q = 0$,
and decreases with $Q$.
%
According to Eq. \eqref{eq:q_opt},
the ordinate of the distribution curve
gives $1/(T\alpha) = 1/(T \dot q)$,
and the area under the curve in the region $Q' < Q$
gives $t/T$.
%
From this picture, we can draw several conclusions.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  }
  \caption{
    \label{fig:massq}
    Geometric characterization of the optimal schedule,
    $\alpha(t)$,
    via the normalized mass distribution, $m(Q)$.
    %
    \note{The figure was produced by \texttt{doc/fig/massq.gp}.
    }%
  }
\end{center}
\end{figure}

For the asymptotic error, we have,
from Eqs.
\eqref{eq:error_asym_Lagrangian}
and
\eqref{eq:Lagrangian_const},
%
\begin{equation}
  E_A = c^2 \, T
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  .
\label{eq:error_asym2}
\end{equation}
%
The optimal value of $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, \alpha_0 / 2 }
  ,
\label{eq:opt_qT}
\end{equation}
i.e., the initial updating magnitude
should be half of $\alpha_0$.

\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial q(T)
  }
  &=
  -\alpha_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial q(T)
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial q(T) = 0$.
}

@@

  \node[draw, text width={width("sampling")}]
    (sys) at (0, 1) {System\\and\\sampling\\process};

  \node[draw, text width={width("Eigenvalues,")}]
    (lambda) at (2.2, 3) {Eigenvalues,\\$\lambda_2, \dots, \lambda_n$};

  \node[draw, text width={width("Integrals, ")}]
    (gamma) at (2.2, 1) {Integrals, \\$\Gamma_2, \dots, \Gamma_n$};

  \node[draw, text width={width("Simulation")}]
    (T) at (2.2, -0.5) {Simulation\\time, $T$};

  \node[draw, text width={width("Optimal")}]
    (qT) at (4.7, 1) {Optimal\\ $q(T)$};

  \node[draw, text width={width("Schedule,")}]
    (alpha) at (7, 2) {Optimal schedule,\\ $\alpha(t)$};

  \node[draw, text width={width("Error, $\Err$")}]
    (err) at (7, 0) {Error, $\Err$};

  \node[inner sep=0, minimum size=0] (M1) at (3.4, 1) {};

  \node[inner sep=0, minimum size=0] (N1) at (5.9, 1) {};

  %\node[inner sep=0, minimum size=0] (R1) at (3.6, 1) {};
  %\node[inner sep=0, minimum size=0] (R2) at (4.6, 0) {};
  %\node[inner sep=0, minimum size=0] (R3) at (5.7, 1) {};

  \draw[->, >=latex, thick]
    (w) edge (lambda)
    (w) edge (gamma)
    (sys) edge (gamma);

  \draw[->, >=latex, thick]
    (lambda) -| (M1)
    (gamma)  -- (M1)
    (T)      -| (M1)
    (M1)     -- (qT);

  \draw[->, >=latex, thick]
    (qT) -- (N1)
    (N1) |- (alpha);

  \draw[->, >=latex, thick]
    (N1) |- (err);

  \draw[->, >=latex, thick]
    (M1) edge[bend left=70] (N1);
    %(R1) |- (R2) -| (R3);
\end{tikzpicture}
\caption{
  \label{fig:vardep}
  Dependence of variables in computing
  the optimal schedule and the error.
}
\end{figure}

@@

\item
Adjust $q(T)$,
to satisfy Eq. \eqref{eq:opt_qT}.

@@

\item
Compute the error $E$ from
Eqs. \eqref{eq:error_split},
  \eqref{eq:error_res1},
  and
  \eqref{eq:error_asym2}.
